---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ShinyLipids

<!-- badges: start -->
![Travis build status](https://travis-ci.com/jannikbuhr/ShinyLipids.svg?token=czApY9arsWuqNrp7yAYj&branch=master)
<!-- badges: end -->

A remake of ShinyLipids with a focus on consistent coding,
documentation, improved functionality and above all, user-friendliness.

## What is ShinyLipids?

ShinyLipids is a visualization tool for lipidomics data built in R with the `shiny` package. The original
version was developed in 2015 by _Mathias Gerl_ and further enhanced by _Manuel Haußmann_ and _Sebastian Bender_ until August, 2017.
It serves as the primary visualization dashboard for the [Brügger Lipidomics Group](https://bzh.db-engine.de/default.asp?lfn=2241&fg=4289) at the Heidelberg Biochemistry Center (BZH).
There is also a great deal of back end infrastructure, built by Mathias, to support a server with a database,
upload tools and a version of this App for internal use at the BZH.

## What changed?

In 2018 I started working on a new version of ShinyLipids.
The idea is still very much based on the original,
but the internal coding and data structure is designed
to be more consistent and host a safer environment
for the implementation of new features within the R code.
More information can be found on [my website](https://jmbuhr.de/project/shinylipids/).

## How do I use ShinyLipids?

Make sure you have the latest version of R and RStudio installed.+

If you are new to R, just head over to

[https://cran.r-project.org/](https://cran.r-project.org/)
and follow the download link suitable for your operating system.
Then, head over to [https://rstudio.com/products/rstudio/download/#download](https://rstudio.com/products/rstudio/download/#download) and install RStudio. You can think
of R as the engine that runs the Code for ShinyLipids and of RStudio
as a car with a nice dashboard around the engine to help you the raw power.
Don't worry, there won't be any coding necessary to get started with your
data in ShinyLipids.

Then run this piece of R code to install ShinyLipids with all it's dependencies^[If you want to publish the Shiny App on RStudio-connect, you will need a couple of additional packages that
are listed in `app.R`.] from github:

```{r, eval=FALSE}
if ( !("devtools" %in% installed.packages()) ) install.packages("devtools")
devtools::install_github("jannikbuhr/ShinyLipids")
```

(To run the code, copy and paste it into the R console that pops up in the
bottom right corner when you first open RStudio. Then hit Enter.)

ShinyLipids gets the data it needs from a database.
This database needs to contain at least the two tables **id_info** and **data2** ^[the naming is due to back end compatibility].
Of course, you might not have your data in a database-dump file. To
get your tabular data (preferably _.csv_) into such a file, you can use a little helper
function in ShinyLipids, which you can run from an R script.
Before you do so, make your your tables contain the columns
listed in section [What should my tables look like?](#tables).

```{r, eval=FALSE}
if ( !("readr" %in% installed.packages()) ) install.packages("readr")
myMedatdata <- readr::read_csv("myMetadata.csv")
myDataset   <- readr::read_csv("myDataset.csv")
ShinyLipids::createDatabase("./inst/extdata/exampleDatabase.db", myMedatdata, myDataset)
```

Now you are ready to look at your lipids! Run this code for a quick start:

```{r, eval=FALSE}
databaseConnection <- DBI::dbConnect(RSQLite::SQLite(), "./inst/extdata/exampleDatabase.db")
ShinyLipids::run_app(db = databaseConnection)
```

### What should my tables look like? {#tables}

**id_info**

This is `myMetadata.csv`. Not all columns are needed.

|Column           | Description                   |
|:----------------| -----------                   |
|id               | whole number:                 |
|title            | text: title of dataset        |
|date_upload      | text: YYYY-MM-DD              |
|status           | text: e.g. in progress        |
|sample_from      | text: owner of sample         |
|date_sample      | text: YYYY-MM-DD              |
|extracted_by     | text: name                    |
|date_extraction  | text: YYYY-MM-DD              |
|measured_by      | text: name                    |
|date_measured    | text: YYYY-MM-DD              |
|distinct_samples | whole number: n of samples    |
|data_lines       | whole number: n of datapoints |
|file             | text: name of raw datafile    |
|instruments      | text: instruments             |

**data2**

This is `myDataset.csv`.

|Column                     | Description                   |
|:--------------------------| -----------                   |
|id                         | whole number: matches metadata|
|sample_identifier          | text: e.g. "1Aa"|
|lipid                      | text: format: "Cer 34:1;2"|
|category                   | text: e.g. SP|
|func_cat                   | text: e.g. SP|
|class                      | text: e.g. SP Cer|
|length                     | whole number: 34|
|db                         | whole number: n of double bounds|
|oh                         | whole number: n of OH groups|
|chains                     | text: format: "34:1;2"|
|chain_sums                 | text: format: "34:1;2"|
|sample                     | text: e.g. "1 - 0min Stim"|
|sample_replicate           | text: e.g. "1 - 0min Stim_A"|
|sample_replicate_technical | text: e.g. "1 - 0min Stim_A_a"|
|value                      | number|

Optionally, your database can contain a table named "LIPID_CLASS_ORDER_COMPLETE" with
a numeric _class_order_ column associated with a _class_ column.
You can also reorder your lipid classes in the app on bottom of the main plotting page.

## For advanced use

If you want to run ShinyLipids on your own server connected to a database
or post your findings to shinyapps.io, you will have to download the source code:

```{bash, eval=FALSE}
git clone https://github.com/jannikbuhr/ShinyLipids.git
```

Open the file `app.R`, read the comments and uncomment the specified lines. Then you
will be able to run the app or use the buttons in the top right corner of your RStudio code panel.

## Notes

Please note that the 'ShinyLipids' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

