---
title: "Data processing in ShinyLipids"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# So what happens to my data?...

you might rightfully ask yourself when loading a dataset into ShinyLipids.
Luckily, we expose the internals of ShinyLipids in the form
of clearly named functions for you to use even outside of
the app. If you want to learn more about _how exactly_ things
work underneath the hood, you can click at the
individual functions to read their documentation or
have a look at the source code.
But for now, let's explore the example data together.
We will follow the steps that ShinyLipids takes and inspect intermediate
results.

First, we load ShinyLipids

```{r setup}
library(ShinyLipids)
```

We can get the file path to our example data, given that
ShinyLipids is installed, and open up a database connection to
the database dump file

```{r}
path <- system.file("extdata/exampleDatabase.db", package = "ShinyLipids")
databaseConnection <- DBI::dbConnect(RSQLite::SQLite(), path)
```

We see, that our example file contains two tables (their names are mandatory).

```{r}
DBI::dbListTables(databaseConnection)
```

From this connection, we load in our meta data "id_info" -- information
about the dataset(s) in the "data2" table:

```{r}
metaData <- collectMetaData(databaseConnection)
metaData
```

We see, that our ownly dataset in the database has id `1`, so
we  use that id to collect the raw data and inspect the first 6 rows:

```{r}
rawData <- collectRawData(con = databaseConnection,
                          id = 1)
head(rawData)
```

Next, we run the data processing steps.

```{r}
plotData <- rawData %>%
  imputeMissingIf() %>% 
  addLipidProperties() %>% 
  standardizeWithinTechnicalReplicatesIf(TRUE) %>%
  filterRawDataFor() %>%
  standardizeRawDataWithin() %>%
  createPlotData()

head(plotData)
```

```{r}
meanPlotData <- summarisePlotData(plotData)
head(meanPlotData)
```


```{r}
plt <- createMainPlot(plotData            = plotData,
                      meanPlotData        = meanPlotData,
                      pairwiseComparisons = NULL)

plt
```

