---
title: "Technical aspects"
description: >
  Learn how ShinyLipids works behind the scenes.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical aspects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Concepts that make ShinyLipids safer

The main reason people switch from spreadsheet software
such as Microsoft Excel to scripting languages such as
Python and R is reproducibility, re-usability and the idea
that code is an important means of communication.

Programming languages offer immense power.
But with great power comes great responsibility.
While analysis in code are less error prone to copy-and-paste errors
as found in Excel and Co., or involuntary conversions
to dates^[One study found that one 5th of screened genetics papers contained errors due to this behaviour in Excel. [Source](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1044-7)],
they can of course contain mistakes nonetheless.

A number of measures have been taken to ensure
that the code presented here is safe, testable and
easy to extend.

## Functional programming

R at it's heart is a functional programming language.
Loosely speaking, this means that functions are first
class citizens in this language. They can, for example,
be passed to other functions (higher order functions)
as arguments just like variables
However, it is not a strict functional language and
functions are not required to be _pure_, that
is free of side-effects. A pure function acts like
a mathematical function: When given the same input,
it produces the same output and does not rely
on any external, global state. I designed the functions
that process your data in ShinyLipids to be pure.
That means, if you put in the same data, rain of shine you
get the same processed data out.
Furthermore, this can guarantee that if some other part of ShinyLipids
changes or is being worked on, the way the pure functions
operate doesn't change.
Understanding and verifying one function does not require
you to read and understand the whole source code of all
other functions.
Of course, mistakes can still happen,
but the more the code aligns with conventions
of pure functional programming, the easier they are to
find and prevent.
But what if we do want to change things?
How can we make sure that `ShinyLipids` still
works as a whole?
This is where unit tests come into play.

## Unit tests

Because the functions mentioned above only rely on their
input, we can also test and verify them
as independent units.
Furthermore, we can automate the whole process!
Every time a code change is pushed online, a whole
bunch of tests are run to make sure that the app
still works, automatically.

# How data flows in ShinyLipids

```{r flowchart, eval=TRUE, echo=FALSE}
DiagrammeR::grViz(diagram = "./dataflow.dot")
```