---
title: "WIP blogpost"
description: >
  Learn ...
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WIP blogpost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document provides a quick outline of the internal data structure of the new Shiny Lipids as well as some thoughts
behind the UI-Changes.

## User Interface

The user interface is based on the popular R-package `shinydashboard`. It features:

- Header
- Sidebar
- Contains most of the control elements that affect all outputs
- Divided into tabs: _Datasets_ opens the datasets-tab in the body,
_Visualization_ hosts the tabs for the main plot, a principal component analysis (PCA) and a heatmap
- Body
- contains the main elements. Plots are always at the top of each panel with additional options below.

## Usage

### Filtering

If you just loaded your dataset into ShinyLipids^[following the
[readme on github](https://github.com/jannikbuhr/ShinyLipids)] (or received a link to the app on shinyapps.io), you will
see the content of the tab _Database Info_. This is your central hub for downloading your raw or filtered dataset
in the standardized comma-separated-values (csv) format^[German _excel_-users beware: German csv-files
are actually semicolon-separated due to the German use of "," as the decimal separator and your excel installation might
expect those while the data you download here is proper _comma_-separated]. If already know that you only want to
look at a subset of your data, have a look in the sidebar and filter your data with the _Filters_ and _Samples_ tabs.
These tabs can of course be accessed while looking at your graphs as well.

### Plotting | Mapping

Now have a look at the most exciting panels: The main plot, PCA and heatmap, coupled with the tab _Mapping_.
"Mapping" refers to a term from the grammar of graphics originally invented by Leland Wilkinson [@Wilkinson.2005] and
adapted for plotting in the R-programming-language by Hadley Wickam [@Wickham.2010]. The term refers to the way
that features of your data (e.g. the length of a lipid, its class, its hydroxylation state etc.) are mapped onto
aesthetic features of your plot (e.g. the x-/y-position, the color, size etc.). Apart from that there is
also the option to separate different slices of your data into windows -- so called "facets" -- , standardize the sum of 
all values within a specific feature to 100% (e.g.) and subtract the values of one sample from all others as a baseline.

The only aesthetic mapping fixed in place is the measured value: it is always mapped to the y-axis for the main plot or 
the color value in the heatmap. Apart from that, you are free to explore different visualizations. Try for example,
to select a subset of your samples in the _Samples_ tab, then map e.g. the class to the x-axis, the functional category
to the color and facet by samples to see them side by side.

Once you are happy with the graph, export it to pdf with the buttons below and safe the underlying data as .csv to
revisit it at a later time or do your own calculations.

### Obvious changes to the old ShinyLipids

Most changes happened under the hood and you can read about those below, but for you as a user there where also a couple
of major changes:

For the new workflow to get the plot you desire, read paragraph [Plotting | Mapping](#plotting-mapping-1).

If you were
a user of the old ShinyLipids, you may miss the color gradient for the heatmap. It has been deprecated in favor
of the [viridis colormap](https://bids.github.io/colormap/). To understand, why the choice of color for heat maps is
important and why the viridis color map is better suited for scientific visualizations,
I highly recommend [this video](https://www.youtube.com/watch?v=xAoljeRJ3lU) by the original authors.

The original version also enabled the user to perform a t-test if a subset of two samples was selected. This feature
was omitted on purpose in the new version. For a number of reasons, a t-test was not the optimal choice. I consider
ShinyLipids a tool for exploratory data analysis, but if you do wish to perform a test for statistical significance,
here are a few things to consider when working with the downloaded data:

- The t-test is a parametric test and assumes your data is (roughly) normal distributed, which will apply to most lipidomics data only after _log-transformation_.
- Looking at your data is a form of comparison! Even if you then perform only one t-test, you need to correct
the p-value for [multiple comparisons](https://www.statisticsdonewrong.com/p-value.html#multiple-comparisons), unless
you clearly stated a hypothesis before the fact.

## The internals

Shiny Lipid _alpha_ makes full use of `shiny`'s reactive functions. Those keep track of their status and only update if
they find themselves outdated.
The overwriting of global variables from inside observer-functions as it was the case in previous versions was abandoned
for the sake of a cleaner data-structure (and more elegant code).

## global.R

Only a handful of global variables and functions are used:

- `database_connection` points to the local `src_sqlite("database/Sqlite.db")` database but could be changed to point
to a database on the server
- `features` is a list of features present in the data and is used to populate the choices for selectInputs in _ui.R_
- SQL-Queries to retrieve the list of data sets and one specific dataset from the database
- Definitions for a `mainTheme` and `mainScale(<number of colors>)` to add to ggplots
- the helper function `is.discrete()` borrowed from the `plyr`-package

## ui.R

Mostly explained in [User Interface](#User-Interface). For more documentation see comments in the code.

## server.R

This is where the magic happens.

## Reactive dependencies

The flowchart ^\@ref(fig:flowchart)^ below was created with [mermaid.js](https://mermaidjs.github.io/).

```{r flowchart, eval=TRUE}
library(DiagrammeR)
mermaid("
graph TB
subgraph Datasets
M[metaData] -->|output| MT(metaDataTable)
MT-->|update selectInput|ID(ID)
MT -->saveMeta(( saveMeta))
end
subgraph mainData
ID-->|select ID|RD(rawData)
RD-->SR((saveRawCSV))
RD -->|stanndardization and filtering|MD(mainData)
MD-->saveMainCSV((saveMainCSV))
FI{filter_/sample_}--> |filtering| MD
MD-->|output|mDT((mainDataTable))
end
subgraph main Plot
MD-->|summarize|PD(plotData)
AES{input$aes_}-->|define|PD
PD-->|average replicates|mPD(meanPlotData)
AES{input$aes_}-->|define|mP
PD-->|ggplot|mP(mainPlt)
mPD-->|ggplot|mP(mainPlt)
mP-->|outout|mPL((mainPlot))
mPL-->|click/brush|ranges
ranges-->|update|mP
end
subgraph PCA
PD-->|spread to matrix|pcaDATA
AES-->|define|pcaDATA
pcaDATA -->|calculate|pcaObject
pcaObject -->|output summary|PCAI((pca_info))
pcaObject -->|ggplot|pca_scores
pcaObject -->|ggplot|pca_loadings
pca_scores -->|render plot|PCAS((pca_scores))
pca_loadings -->|render plot|PCAL((pca_loadings))
end
subgraph heatmap
AES-->|define|HP(heatPlt)
PD-->|ggplot|HP
HP-->|output|HPO((heatPlot))
end
")
```

