---
title: Shiny Lipids
description: Getting Started with Shiny Lipids
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Getting your feet wet with Shiny Lipids

### Installation 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

This is how you install Shiny Lipids

```{r two-plus-two, exercise=TRUE}

```

### Loading example data

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Exploring your data outside of shiny

### Main functions of Shiny Lipids

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

